Limits to Ecological Forecasting: Estimating Uncertainty for Critical Transitions with Deep Learning

### 2022-08-14 manuscript/manuscript.Rmd
Added explicit definitions of scores

header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{lineno}
  - \usepackage{setspace}
  - \linenumbers
  - \doublespacing

@@

```{r figure1, fig.cap="Forecast scenarios. A. The Hopf bifurcation: a stable node develops into a limit cycle which gradually grows larger in this predator-prey model. B. The saddle-node bifurcation in a single species. C. The stochastic transition in a single species.  Plots show historical data used to train the algorithm in purple, and replicate simulations of the true dynamics ('future ensemble') in yellow. Note how the characteristic time for the critical transition varies across the transitions. We will examine various models which will each produce probablistic forecast distributions (blue, Fig 3-5) seeking to match the true future ensemble (yellow) as closely as possible.", fig.height=6}
hopf <- read_csv("../greta/forecasts/simulation=hopf/hopf_increasing.csv.gz")
panelA <- hopf |> 
  mutate(type = as.factor(type)) |>
  filter(type %in% c("historical", "true"), i < 10) |>
  rename(species = variable) |> 
  mutate(species = forcats::fct_recode(species, host = "H", parasitoid = "P")) |>
  bind_rows(tibble(i=1,t=1,type="predicted", variable="H", value=NA)) |> #placeholder
  mutate(type = forcats::fct_recode(type, "future ensemble" = "true", "forecast" = "predicted")) |>
  ggplot(aes(t, value, group=interaction(species, i, type), col=type, lty=species)) + 
  geom_line(show.legend = TRUE, alpha = 0.7) +
  theme_bw() +  
  scale_color_viridis_d(guide="none") +
  ggtitle("A. Hopf bifurcation")+
  theme(legend.position="top")

sn <- read_csv("../greta/forecasts/simulation=saddle/saddlenode.csv.gz")
panelB <- sn |>
  mutate(type = as.factor(type)) |>
  filter(type == "historical", i == 1) |>
  bind_rows( filter(sn, type=="true")) |>
  bind_rows(tibble(i=1,t=1,type="predicted", variable="N", value=NA)) |> #placeholder
  mutate(type = forcats::fct_recode(type, "future ensemble" = "true", "forecast" = "predicted")) |>
  ggplot(aes(t, value, col = type, group=interaction(i,type), lty=type)) +
  geom_path(alpha=0.5, show.legend = FALSE) +
  scale_color_viridis_d() +
  theme_bw() + 
  ggtitle("B. Saddle-node bifurcation")

stoch <- read_csv("../greta/forecasts/simulation=stochastic/stochastic.csv.gz")
panelC <- stoch |>
  filter(type == "historical", i == 1) |>
  bind_rows(filter(stoch, type=="true", i < 50)) |> 
  bind_rows(tibble(i=1,t=1,type="predicted", variable="N", value=NA)) |> #placeholder
  mutate(type = forcats::fct_recode(type, "future ensemble" = "true", "forecast" = "predicted")) |>
  ggplot(aes(t, value, col = type, group=interaction(i,type), lty=type)) +
  geom_path(alpha=0.5) + 
  scale_color_viridis_d() +
#  scale_color_grey(end=0.6) + 
  theme_bw() + 
  ggtitle("C. Stochastic transition") +
  theme(legend.position="bottom")

@@

To compare forecasts, we focus exclusively on metrics of forecast skill which satisfy the property from @gneiting_strictly_2007 of a strictly proper score.
This ensures the very desirable behavior that no probabilistic forecast $Q(x,t)$ can have a score as high as the score of the true process $P(x,t)$ on average. 
In other words, while it is possible for any of the models considered to _overfit_ the data against which they are trained, i.e. have a higher likelihood than the true process, it is not possible for these models to overfit the data against which they are scored.
It is worth noting that this property applies specifically to probabilistic forecasts and not point forecasts. Not all common metrics often used to compare forecasts are strictly proper -- such as the average root-mean-square error or the average absolute error. <!-- MFL: hmm I am not completely following this point here. CB: added some examples of non-strictly-proper scoring metrics. For instance, {darts} has a whole 'metrics' class, which does not include any strictly proper metrics (because ML researchers don't care about statistical theorems?) -->
Concerns about over-fitting arise in most types of model estimation and are a particularly acute concern to machine learning methods due to the bias-variance trade-off [@mehta_high-bias_2019]. <!-- MFL: I don't think I nailed the overfitting part in ML section but we probably want to reference this here-->
This makes the use of strictly proper scoring especially relevant in assessing machine learning predictions.

Not even all strictly proper scores will agree on the same relative ranking between forecasts.
We will focus on two of the most common such skill metrics, CRPS score and log probability score (negative log likelihood) [e.g. see @gneiting_strictly_2007; @gneiting_probabilistic_2014].
Of the two, the logs probability score puts a much a greater penalty on unexpected observations than CRPS, and may be more suitable when the occurrence of unexpected events incurs a particularly high cost.
Note that while the minus log-likelihood can be negative for sufficiently high probability densities, we use a fixed scalar shift of logs score to ensure the log skill score is strictly positive, which facilitates visualization without impacting relative rankings.

@@

```{r figure2, dev="png", dpi=300, dev.args = list(type = "cairo-png"), out.width="400px", out.height="300px"}
#| fig.cap="Overall distribution of skill scores across models, including an ensemble of methods.
#|  Smaller scores are better (indicating smaller errors).
#|  Black bars indicate means, dots indicate all individual predictions over time and replicate 'true'
#| simulations of the given scenario."
shift |> 
  ggplot(aes(forecasting_model, score, color=forecasting_model)) + 
  geom_jitter(alpha = 0.06, show.legend = FALSE, shape=".") + 
  stat_summary(fun = mean, shape="|", show.legend = FALSE, col="black") +
  scale_x_discrete(limits = rev) +
  coord_flip() + 
  facet_grid(simulation ~ metric, scales = "free") + 
  scale_y_log10() +
  theme_bw() +
  theme(axis.title.y = element_blank())
```

@@

```{r}
forecasts <- read_parquet("data/forecasts.parquet") |>
  filter(forecasting_model != "ml_ensemble") |> 
  bind_rows(tibble(type= c("future ensemble", "historical"))) |>
  mutate(type = forcats::fct_recode(type, "future ensemble" = "observed", "forecast" = "predicted"))

observations <- read_parquet("data/observations.parquet") |> 
  mutate(type = forcats::fct_recode(type, "future ensemble" = "observed", "forecast" = "predicted"))

@@

```{r figure3}
#| fig.cap="Forecasts of the Hopf bifurcation under each model, compared to 15 realizations of the true model. 
#|  The bifurcation occurs soon after forecasting period begins, 
#| leading to progressively larger and larger oscillations. 
#|  Prior to the bifurcation, pseudo-cycles are visible in the training data due to stochastic excitations. 
#|  Following the bifurcation, stochasticity blurs the oscillatory pattern across replicate simulations. 
#|  Only last 25 time points of training data are shown."

forecasts  |> 
  filter(iter %in% 1:15, simulation == "hopf", variable !="X") |> 
  bind_rows(tibble(type= c("future ensemble", "historical"), forecasting_model="MCMC", variable="host")) |>
  mutate(type = factor(type,levels=c("historical","forecast", "future ensemble"))) |>
  ggplot(aes(x=t, y=value, group = interaction(iter, variable), col=type)) + 
  geom_line(alpha=0.1) +
  geom_line(alpha=0.1, data = hopf_obs) +
  geom_line(data = filter(hopf_obs, type=="historical")) +
  facet_grid(forecasting_model ~ variable, scales = "free") +
  theme_bw()  + 
  scale_color_viridis_d() +
  coord_cartesian(ylim = c(0,15), xlim=c(75, 200)) + 
  ggtitle("Hopf bifurcation")
```

@@

```{r figure4}
#| fig.cap = "Forecasts of the saddle-node bifurcation under each model, compared to 15 realizations of the true model. 
#|  Training data proceeds the bifurcation, making accurate prediction without knowledge of the underlying model very difficult. "

forecasts |> 
  filter(iter %in% 1:15, simulation == "saddle") |> 
  bind_rows(tibble(type= c("future ensemble", "historical"), forecasting_model="MCMC", variable="host")) |>
  mutate(type = factor(type,levels=c("historical","forecast", "future ensemble"))) |> 
  ggplot(aes(x=t, y=value, group = interaction(iter, variable), col=type)) + 
  geom_line(alpha=0.1) +
  geom_line(alpha=0.1, data = saddle_obs) +
  geom_line(data = filter(saddle_obs, type=="historical")) +
  facet_wrap(~forecasting_model, scales = "free") +
  theme_bw()  + 
  scale_color_viridis_d() +
  coord_cartesian(ylim = c(0,1)) + 
  ggtitle("Saddle-Node bifurcation")
```

@@

```{r figure5}
#| fig.cap = "Forecasts of the stochastic transition under each model, compared to 15 realizations of the true model. 
#|  In contrast to the other challenges, this case considers the prediction of replicate systems starting from the same initial condition, rather than forecasting the future evolution of the model after the stochastic transition has already occurred."
forecasts |>
  filter(iter %in% 1:15, simulation == "stochastic") |> 
  bind_rows(tibble(type= c("future ensemble", "historical"), forecasting_model="MCMC", variable="host")) |>
  mutate(type = factor(type,levels=c("historical","forecast", "future ensemble"))) |> 
  ggplot(aes(x=t, y=value, group = interaction(iter, variable), col=type)) + 
  geom_line(alpha=0.2) +
  geom_line(alpha=0.2, data = filter(stochastic_obs, type=="future ensemble")) +
  geom_line(data = filter(stochastic_obs, type=="historical")) +
  facet_wrap(~forecasting_model, scales = "free") +
  theme_bw()  + 
  scale_color_viridis_d() +
  coord_cartesian(ylim = c(0,0.8)) + 
  ggtitle("Stochastic transition")
```

@@

Examining patterns in the scores over time (Fig 6-7) provides a more nuanced understanding of the forecast dynamics than aggregate scores alone (Fig 2).
In the Hopf bifurcation, CRPS scores get worse over time across all methods, including the MCMC forecasts.
In the saddle node bifurcation and stochastic transition, the same pattern holds somewhat more dramatically for non-MCMC forecast,
while MCMC scores are worst in mid-range.
Comparing CRPS scores to logs score also emphasizes the relative role of uncertainty: for instance, the MCMC scores for the Hopf bifurcation get steadily worse under CRPS but not under logs score. 
A relatively sharp transition can be seen under both MCMC scores on the Hopf bifurcation once the magnitude of the oscillations exceeds the variance created by mere stochasticity: the MCMC model no longer over-estimates the spread of the data, while the ML models now underestimate that variation. 
CRPS scores for stochastic transitions exhibit a distinct two-branch pattern, with scores for a given replicate being either very high (poor skill) or very low, reflecting whether the individual 'true' replicate matches the mean state predicted by the forecast or the other state. 
Logs skill score may be a better measure in this context, where correctly capturing the uncertainty in the forecast means that this bi-modal structure in scores can be avoided entirely, e.g. by the MCMC predictions. 
The forecast-skill-over time plots illustrate different reasons for the bi-modal distribution in skill seen for the saddle-node and stochastic transition scenarios in Figure 2 respectively: in the case of the saddle node, the two modes are distinguished by time-horizon: short term forecasts are relatively accurate, longer term forecasts (i.e. after the catastrophic transition) are poor.
In the case of the stochastic transition model, the two modes are not structured by horizon but by replicate, with some replicates having transitioned and others still in the original state.

```{r figure6}
#| fig.cap="CRPS scores over time for each scenario.
#|  Each line represents the scores against a replicate of time series observations from 'true model'. 
#|  Patterns show forecast skill generally get worse over time -- gradually 
#| in the case of Hopf bifurcation or suddenly in response to the saddle-node bifurcation. 
#|  The MCMC shows a sharp improvement in forecast skill over time as it passes the transition window. 
#|  In the stochastic transition context, this creates two branches, where high values indicate periods of 
#| time when the forecast predicts the wrong state and lower branch indicates the correct state."

@@

```{r figure7}
#| fig.cap="Logs skill score over time. 
#|  Forecasts which underestimate uncertainty do substantially worse in logs score than in CRPS score. 
#|  Comparing this panel to those in Figure 6 highlights scenarios that most often underestimate uncertainty. 
#|  Generally, MCMC performs better relative to other models under this metric than it does under CRPS, 
#| reflecting the bias-variance tradeoff of machine learning."

>>>>

header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{lineno}
  - \usepackage{setspace}
  - \usepackage{bbm}
  - \linenumbers
  - \doublespacing

@@

```{r fig.cap="Forecast scenarios. A. The Hopf bifurcation: a stable node develops into a limit cycle. B. The saddle-node bifurcation. C. The stochastic transition."}
hopf <- read_csv("../greta/forecasts/simulation=hopf/hopf_increasing.csv.gz")
panelA <- hopf |> filter(type %in% c("historical", "true"), i < 10) |>
  rename(species = variable) |> 
  mutate(species = forcats::fct_recode(species, host = "H", parasitoid = "P")) |>
  ggplot(aes(t, value, group=interaction(species, i, type), col=species)) + 
  geom_line(show.legend = FALSE, alpha = 0.7) + theme_bw()+  scale_color_grey() +
  ggtitle("A. Hopf bifurcation")

sn <- read_csv("../greta/forecasts/simulation=saddle/saddlenode.csv.gz")
panelB <- sn |> filter(type == "historical", i == 1) |>
  bind_rows(filter(sn, type=="true")|>mutate(type="predicted")) |>
  ggplot(aes(t, value, col = type, group=interaction(i,type), lty=type)) +
  geom_path(alpha=0.5, show.legend = FALSE) +
  scale_color_grey(end=0.6) + theme_bw() + 
  ggtitle("B. Saddle-node bifurcation")

stoch <- read_csv("../greta/forecasts/simulation=stochastic/stochastic.csv.gz")
panelC <- stoch |> filter(type == "historical", i == 1) |>
  bind_rows(filter(stoch, type=="true", i < 50) |> mutate(type="predicted")) |>
  ggplot(aes(t, value, col = type, group=interaction(i,type), lty=type)) +
  geom_path(alpha=0.5) + scale_color_grey(end=0.6) + theme_bw() + 
  ggtitle("C. Stochastic transition")

@@

To compare forecasts, we focus exclusively on metrics of forecast skill which satisfy the property from @gneiting_strictly_2007 of a strictly proper score.
This ensures the very desirable behavior that no probabilistic forecast $Q(x,t)$ can have a score as high as the score of the true process $P(x,t)$ on average. 
In other words, while it is possible for any of the models considered to _overfit_ the data against which they are trained, i.e. have a higher likelihood than the true process, it is not possible for these models to overfit the data against which they are scored.
It is worth noting that this property applies specifically to probabilistic forecasts and not point forecasts. Not all common metrics often used to compare forecasts are strictly proper -- such as the average root-mean-square error or the average absolute error.
Concerns about over-fitting arise in most types of model estimation and are a particularly acute concern to machine learning methods due to the bias-variance trade-off [@mehta_high-bias_2019].
This makes the use of strictly proper scoring especially relevant in assessing machine learning predictions.

Not even all strictly proper scores will agree on the same relative ranking between forecasts.
We will focus on two of the most common such skill metrics, CRPS score and log probability score (negative log likelihood) [e.g. see @gneiting_strictly_2007; @gneiting_probabilistic_2014].
We have defined these two scores explicitly in Equations 7-8, where $F$ and $f$ are the respective CDF's and PDF's from a probabilistic forecast, and $y$ refers to an observation.
Of the two, the logs probability score puts a much a greater penalty on unexpected observations than CRPS, and may be more suitable when the occurrence of unexpected events incurs a particularly high cost.
Note that while the minus log-likelihood can be negative for sufficiently high probability densities, we use a fixed scalar shift of logs score to ensure the log skill score is strictly positive, which facilitates visualization without impacting relative rankings.

\begin{align} 
\text{LogS}(F, y) &= -\log{f(y)} \\
\text{CRPS}(F, y) &= -\int (F(z) - \mathbbm{1}\{y \leq z\})^2 \, dz
\end{align}

@@

```{r}
#| fig.cap="Overall distribution of skill scores across models, including an ensemble of methods.
#|  Smaller scores are better (indicating smaller errors).
#|  Black bars indicate means, dots indicate all individual predictions over time and replicate 'true'
#| simulations of the given scenario."
shift |> 
  ggplot(aes(forecasting_model, score, color=forecasting_model)) + 
  geom_jitter(alpha = 0.06, show.legend = FALSE, shape=".") + 
  stat_summary(fun = mean, shape="|", show.legend = FALSE, col="black") +
  coord_flip() +
  facet_grid(simulation ~ metric, scales = "free") + 
  scale_y_log10() +
  theme_bw() +
  theme(axis.title.y = element_blank())
```

@@

```{r}
forecasts <- read_parquet("data/forecasts.parquet") |>
  filter(forecasting_model != "ml_ensemble")
observations <- read_parquet("data/observations.parquet")

@@

```{r}
#| fig.cap="Forecasts of the Hopf bifurcation under each model, compared to 15 realizations of the true model. 
#|  The bifurcation occurs soon after forecasting period begins, 
#| leading to progressively larger and larger oscillations. 
#|  Prior to the bifurcation, pseudo-cycles are visible in the training data due to stochastic excitations. 
#|  Following the bifurcation, stochasticity blurs the oscillatory pattern across replicate simulations. 
#|  Only last 25 time points of training data are shown."

forecasts  |> 
  filter(iter %in% 1:15, simulation == "hopf", variable !="X") |> 
  ggplot(aes(x=t, y=value, group = interaction(iter, variable), col=type)) + 
  geom_line(alpha=0.1, data = hopf_obs) +
  geom_line(alpha=0.1) +
  geom_line(data = filter(hopf_obs, type=="historical")) +
  facet_grid(forecasting_model ~ variable, scales = "free") +
  theme_bw()  + 
  coord_cartesian(ylim = c(0,15), xlim=c(75, 200)) + 
  ggtitle("Hopf bifurcation")
```

@@

```{r }
#| fig.cap = "Forecasts of the saddle-node bifurcation under each model, compared to 15 realizations of the true model. 
#|  Training data proceeds the bifurcation, making accurate prediction without knowledge of the underlying model very difficult. "

forecasts |> 
  filter(iter %in% 1:15, simulation == "saddle") |> 
  ggplot(aes(x=t, y=value, group = interaction(iter, variable), col=type)) + 
  geom_line(alpha=0.1, data = saddle_obs) +
  geom_line(alpha=0.1) +
  geom_line(data = filter(saddle_obs, type=="historical")) +
  facet_wrap(~forecasting_model, scales = "free") +
  theme_bw()  + coord_cartesian(ylim = c(0,1)) + 
  ggtitle("Saddle-Node bifurcation")
```

@@

```{r}
#| fig.cap = "Forecasts of the stochastic transition under each model, compared to 15 realizations of the true model. 
#|  In contrast to the other challenges, this case considers the prediction of replicate systems starting from the same initial condition, rather than forecasting the future evolution of the model after the stochastic transition has already occurred."
forecasts |>
  filter(iter %in% 1:15, simulation == "stochastic") |> 
  ggplot(aes(x=t, y=value, group = interaction(iter, variable), col=type)) + 
  geom_line(alpha=0.2, data = filter(stochastic_obs, type=="observed")) +
  geom_line(alpha=0.2) +
  geom_line(data = filter(stochastic_obs, type=="historical")) +
  facet_wrap(~forecasting_model, scales = "free") +
  theme_bw()  + coord_cartesian(ylim = c(0,0.8)) + 
  ggtitle("Stochastic transition")
```

@@

Examining patterns in the scores over time (Fig 6-7) provides a more nuanced understanding of the forecast dynamics than aggregate scores alone (Fig 2).
In the Hopf bifurcation, CRPS scores get worse over time across all methods, including the MCMC forecasts.
In the saddle node bifurcation and stochastic transition, the same pattern holds somewhat more dramatically for non-MCMC forecast,
while MCMC scores are worst in mid-range.
Comparing CRPS scores to logs score also emphasizes the relative role of uncertainty: for instance, the MCMC scores for the Hopf bifurcation get steadily worse under CRPS but not under logs score. 
A relatively sharp transition can be seen under both MCMC scores on the Hopf bifurcation once the magnitude of the oscillations exceeds the variance created by mere stochasticity: the MCMC model no longer over-estimates the spread of the data, while the ML models now underestimate that variation. 
CRPS scores for stochastic transitions exhibit a distinct two-branch pattern, with scores for a given replicate being either very high (poor skill) or very low, reflecting whether the individual 'true' replicate matches the mean state predicted by the forecast or the other state. 
Logs skill score may be a better measure in this context, where correctly capturing the uncertainty in the forecast means that this bi-modal structure in scores can be avoided entirely, e.g. by the MCMC predictions. 
The forecast-skill-over time plots illustrate diferent reasons for the bi-modal distribution in skill seen for the saddle-node and stochastic transition scenarios in Figure 2 respectively: in the case of the saddle node, the two modes are distinguished by time-horizon: short term forecasts are relatively accurate, longer term forecasts (i.e. after the catastrophic transition) are poor.
In the case of the stochastic transition model, the two modes are not structured by horizon but by replicate, with some replicates having transitioned and others still in the original state.

```{r}
#| fig.cap="CRPS scores over time for each scenario.
#|  Each line represents the scores against a replicate of time series observations from 'true model'. 
#|  Patterns show forecast skill generally get worse over time -- 
#| gradually in the case of Hopf bifurcation or suddenly in response to the saddle-node bifurcation. 
#|  The MCMC shows a sharp improvement in forecast skill over time as it passes the transition window. 
#|  In the stochastic transition context, this creates two branches, where high values indicate periods of 
#| time when the forecast predicts the wrong state and lower branch indicates the correct state."

@@

```{r}
#| fig.cap="Logs skill score over time. 
#|  Forecasts which underestimate uncertainty do substantially worse in logs score than in CRPS score. 
#|  Comparing this panel to those in Figure 6 highlights scenarios that most often underestimate uncertainty. 
#|  Generally, MCMC performs better relative to other models under this metric than it does under CRPS, 
#| reflecting the bias-variance tradeoff of machine learning."
